Exploring probability distributions for bivariate temporal granularities

### 2019-10-30 main.Rmd
interaction section update

When we plot some observations with two circular or aperiodic granularities used as aesthetics or facets, problems arise when we have empty combinations. It is important to check how these two granularities interact with each other, before we expect to gain insights by plotting them together.

@@

Firstly, empty combinations can arise due to the structure of the calendar or hierarchy.
These are called structurally empty combinations. Let us take a specific example, where $C_1$ maps row numbers to Day-of-Month and $C_2$ maps row numbers to Week-of-Month. Here $C_1$ can take 31 values while $C_2$ can take 5 values. There will be $31\times 5=155$ sets $S_{ij}$ corresponding to the possible combinations of WOM and DOM. Many of these are empty. For example $S_{1,5}$, $S_{21,2}$, etc. In fact, most of these 155 sets will be empty, making the combination of $C_1$ and $C_2$ in a graph unhelpful. These are structurally empty sets in that it is impossible for them to have any observations.

Secondly, empty combinations can turn up due to difference in occurance of event location or duration in a calendar. These are called event-driven empty combinations. Again, let us consider a specific example to illustrate this. Let $C_1$ be DOW and $C_2$ be WorkingDay/NonWorkingDay. Here $C_1$ can take 7 values while $C_2$ can take 2 values. So there are 14 sets $S_{ij}$ corresponding to the possible combinations of DOW and WD/NWD. While potentially all of these can be non-empty (it is possible to have a public holiday on any DOW), in practice many of these combinations will probably have very few observations. For example, there are few if any public holidays on Wednesdays or Thursdays in any given year in Melbourne.

@@

Thirdly, empty combinations can be a result of how granularities are constructed. Let $C_1$ maps row numbers to "Business days", which are days from Monday to Friday except holidays and $C_2$ is Day-of-Month. Then the weekends in Days-of-Month would not correspond to any Business days and would have missing observations due to the way the granularities are constructed. this is different from the first case because these are not structurally empty combinations. The structure of the calendar does not lead to these missing combinations, but the construction of the granularity does. Hence, they are referred to as "build-based" empty combinations.

Another example could be where $C_1$ maps row numbers to Day-of-Week and $C_2$ maps row numbers to Month-of-Year. Here $C_1$ can take 7 values while $C_2$ can take 12 values. So there are $12\times7=84$ sets $S_{ij}$ corresponding to the possible combinations of DOW and MOY. All of these are non-empty because every DOW can occur in every month. So graphics involving $C_1$ and $C_2$ are potentially useful.

We investigate some combinations of circular time granularities (periodic/aperiodic) which facilitate or hinder exploratory analysis. The combinations of circular granularities which promote the exploratory analysis through visualization are referred to as **harmonies** and the ones which impede the analysis are referred to as **clashes**.

>>>>

When we plot some observations with two circular or aperiodic granularities used as aesthetics or facets, problems might arise when we have empty combinations. It is important to check how these two granularities interact with each other, before we expect to gain insights by plotting them together.

@@

Firstly, empty combinations can arise due to the structure of the calendar or hierarchy.
These are called "structurally" empty combinations. Let us take a specific example, where $C_1$ maps row numbers to Day-of-Month and $C_2$ maps row numbers to Week-of-Month. Here $C_1$ can take 31 values while $C_2$ can take 5 values. There will be $31\times 5=155$ sets $S_{ij}$ corresponding to the possible combinations of WOM and DOM. Many of these are empty. For example $S_{1,5}$, $S_{21,2}$, etc.This is also intuitive since the first day of the month can never correspond to fifth week of the month. In fact, most of these 155 sets will be empty, making the combination of $C_1$ and $C_2$ in a graph unhelpful. These are structurally empty sets in that it is impossible for them to have any observations.

Secondly, empty combinations can turn up due to differences in event location or duration in a calendar. These are called "event-driven" empty combinations. Again, let us consider a specific example to illustrate this. Let $C_1$ be DOW and $C_2$ be WorkingDay/NonWorkingDay. Here $C_1$ can take 7 values while $C_2$ can take 2 values. So there are 14 sets $S_{ij}$ corresponding to the possible combinations of DOW and WD/NWD. While potentially all of these can be non-empty (it is possible to have a public holiday on any DOW), in practice many of these combinations will probably have very few observations. For example, there are few if any public holidays on Wednesdays or Thursdays in any given year in Melbourne.

@@

Thirdly, empty combinations can be a result of how granularities are constructed. Let $C_1$ maps row numbers to "Business days", which are days from Monday to Friday except holidays and $C_2$ is Day-of-Month. Then the weekends in Days-of-Month would not correspond to any Business days and would have missing observations due to the way the granularities are constructed. This is different from the structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the construction of the granularity does. Hence, they are referred to as "build-based" empty combinations.

An example when there will be no empty combinations could be where $C_1$ maps row numbers to Day-of-Week and $C_2$ maps row numbers to Month-of-Year. Here $C_1$ can take 7 values while $C_2$ can take 12 values. So there are $12\times7=84$ sets $S_{ij}$ corresponding to the possible combinations of DOW and MOY. All of these are non-empty because every DOW can occur in every month. So graphics involving $C_1$ and $C_2$ are potentially useful.

The combinations of circular/aperiodic granularities which lead to structurally, event-driven or build-based empty-combinations are referred to as to as **clashes**.
And the ones that ones that do not lead to any missing combinations are called **harmonies** as they promote the exploratory analysis through visualization.

Even with harmonies, problems might occur due to rarely occuring categories. suppose we have $T$ hourly observations, and $C_1$ is HOD and $C_2$ is DOW. Each element of $C_1$ occurs approximately $T/n_1$ times while each element of $C_2$ occurs approximately $T/n_2$ times. There are no structurally empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(n_1n_2)$. If we require at least $k$ observations to create a meaningful panel, then provided $T \ge n_1n_2k$, the visualization will be acceptable. The value of $k$ will depend on what type of visualization we are producing. For scatterplots, even $k=3$ may be acceptable, but for density estimates, we probably need $k \ge 30$.

The above discussion shows that problems will occur when $n_1n_2 > T/k$, even without structurally empty combinations.

Problems will also occur when we have rarely occurring categories such as the 366th day of the year, or the 31st day of the month.
